---
title: "Japanese Production CDI-CAT"
author: "George"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=F}
library(tidyverse)
library(ggrepel)
library(mirt) # requires package loading, not just referencing. 
library(mirtCAT)
library(here)
library(kableExtra)
library(gridExtra)
library(printr)
library(parallel)
library(doParallel)
RUN_MODELS = FALSE # set TRUE to re-run IRT models (a few hours); otherwise loads saved fits

source("IRT_helpers.R")

language = "Japanese"
form = "WS" # and WG
```


# Introduction

Our goal is to develop and test via simulation a bank of CDI items and IRT parameters that can be used for a CDI-CAT in `r language`.
Our approach is as follows:
We first fit basic IRT models (1-parameter logistic (1PL; i.e. Rasch), 2PL, and 3PL) to CDI data and perform a model comparison.
For the favored model, we then identify candidate items for removal based on low total item information, and then use the (full) item bank in a variety of computerized adaptive test (CAT) simulations on wordbank data.
We provide recommendations for CAT algorithms and stopping rules to be passed on to CAT developers, and benchmark CAT performance compared to random baselines tests of a similar length.

# Data


```{r, load-data, echo=F}
# see 00-preprocessing.R for merging WG + WS datasets 
base::load("data/Japanese-CDI-WS-WG-combined-data-disambig.Rdata")

d_mat <- d_wide %>% 
  data.frame %>%
  select(-id) %>% #  -age, -sex
  data.matrix

too_young <- which(d_demo$age < 12) # 72 children too young to be producing words...
too_old <- which(d_demo$age > 36) # 65 older than the target age range of the test
no_words <- which(d_demo$production==0 & d_demo$age>=12) # 110 additional not producing

to_remove = unique(c(too_young, no_words, too_old))
```

We use the combined production data from `r nrow(d_demo)` participants. 
From this data we remove `r length(too_young)` children <12 months of age, who should not be producing any words yet. 
We also remove an additional `r length(no_words)` children 12+ months of age who are not yet producing any words, as these children cannot be used to fit the IRT models.
Finally, we remove `r length(too_old)` children >36 months of age, as they are beyond the targeted age range of the CDI:WS, and the proposed CAT.
The production sumscores by age for the remaining children are shown below.

```{r, echo=F}
d_demo = d_demo[-to_remove,] # 1160 down to 913
d_mat = d_mat[-to_remove,]
d_demo <- d_demo %>% 
  mutate(age_mos = round(age),
         data_id = paste0(id,"-",age_mos)) # unique-ify ID (some longitudinal data)

rownames(d_mat) = d_demo$data_id

#d_demo$production = rowSums(d_mat[,2:ncol(d_mat)], na.rm=T) 
#table(rowSums(d_mat, na.rm=T) == d_demo$production)

table(d_demo$age_mos) %>% kable(col.names=c("Age","N"))
```


```{r, plot-production, fig.width=6.5, fig.height=4.5, echo=F, warning=F}
d_ag <- d_demo %>% group_by(age_mos) %>%
  summarise(mean=round(mean(production), 2), median=median(production))

ggplot(d_demo, aes(x = age, y = production, color=sex)) + # color=sex
  stat_smooth(method = "lm", formula = y ~ 0 + x + I(x^2), linewidth = 1) +
  geom_jitter(alpha = .2) + theme_bw() + xlab("Age (months)") + ylab("Production Sumscore")
```


# IRT Models

We fit each type of basic IRT model (Rasch, 2PL, and 3PL) using the `mirt` package.

```{r psycho-models_1pl, echo=F, eval = RUN_MODELS}
set.seed(1234)
# problem with item 1 (aïe) - just remove it
mod_1pl <- mirt(d_mat, 1, itemtype='Rasch', verbose=TRUE, technical=list(NCYCLES=4000)) 
# Log-likelihood was decreasing near the ML solution. EM method may be unstable
coefs_1pl <- as_tibble(coef(mod_1pl, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(mod_1pl, simplify = TRUE)$items))
fscores_1pl <- tibble(data_id = rownames(d_mat), 
                      ability = fscores(mod_1pl, method = "MAP")[,1]) 
save(file = here("data/jp_ws_wg_mod_1pl.Rds"), "mod_1pl", "fscores_1pl", "coefs_1pl")
```

```{r, echo=F, eval = RUN_MODELS}
#mod_string = paste0('G = 1-',ncol(d_mat),',
#                LBOUND = (1-',ncol(d_mat),', a1, 0),
#                PRIOR = (1-',ncol(d_mat),', a1, norm, 0.2, 2),
#                PRIOR = (1-',ncol(d_mat),', d, norm, 0, 3)')
mod_string = paste0('G = 1-',ncol(d_mat),',
                LBOUND = (1-',ncol(d_mat),', a1, 0),
                PRIOR = (1-',ncol(d_mat),', d, norm, 0, 3)')
mod <- mirt.model(mod_string)
mod_2pl = mirt(data = d_mat, model = mod, itemtype="2PL", 
                        method="QMCEM", verbose=TRUE, 
                        technical=list(NCYCLES=3000))

coefs_2pl <- as_tibble(coef(mod_2pl, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(mod_2pl, simplify = TRUE)$items))
fscores_2pl <- tibble(data_id = rownames(d_mat), 
                             ability = fscores(mod_2pl, method = "MAP")[,1])
# The following factor score estimates failed to converge successfully: 27,66,323,693
save(file = here("data/jp_ws_wg_mod_2pl_prior.Rds"), "mod_2pl","fscores_2pl", "coefs_2pl")
```


```{r psycho-2pl_coefs, echo=F, eval = RUN_MODELS}
mod_2pl <- mirt(d_mat, 1, itemtype='2PL', verbose=TRUE,  technical=list(NCYCLES=3000))

coefs_2pl <- as_tibble(coef(mod_2pl, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(mod_2pl, simplify = TRUE)$items))
fscores_2pl <- tibble(data_id = rownames(d_mat), 
                             ability = fscores(mod_2pl, method = "MAP")[,1])
# The following factor score estimates failed to converge successfully:  15,242,446
save(file = here("data/jp_ws_wg_mod_2pl.Rds"), "mod_2pl","fscores_2pl", "coefs_2pl")
```

```{r psycho-fit_irt_3pl, echo=F, eval = RUN_MODELS}
mod_3pl <- mirt::mirt(d_mat, 1, itemtype='3PL', verbose=TRUE, 
                      technical=list(NCYCLES=4000))

coefs_3pl <- as_tibble(coef(mod_3pl, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(mod_3pl, simplify = TRUE)$items))
fscores_3pl <- tibble(data_id = rownames(d_mat), 
                             ability = fscores(mod_3pl, method = "MAP")[,1])
# The following factor score estimates failed to converge successfully:
#   8,282,365,824

save(file = here("data/jp_ws_wg_mod_3pl.Rds"), "mod_3pl","fscores_3pl", "coefs_3pl")
```


```{r psycho-models_load, echo=F}
load(here("data/jp_ws_wg_mod_1pl.Rds"))
load(here("data/jp_ws_wg_mod_2pl_prior.Rds"))
#load(here("data/jp_ws_wg_mod_2pl.Rds"))
load(here("data/jp_ws_wg_mod_3pl.Rds"))
```

```{r compare-2pl-with-n-without-prior, include=F, echo=F, eval=F}
# original (no prior) - ends up with some pretty large d and a values
load(here("data/jp_ws_wg_mod_2pl.Rds"))
old_coefs <- coefs_2pl
old_fscores <- fscores_2pl
summary(mod_2pl)
# prop var: .734 

# prior on d ~ norm(0,3) - 
load(here("data/jp_ws_wg_mod_2pl_prior.Rds"))
new_coefs <- coefs_2pl
new_fscores <- fscores_2pl
summary(mod_2pl) 
# prop var: .896 



plot(old_fscores$ability, new_fscores$ability)
cor(old_fscores$ability, new_fscores$ability) # .997
plot(d_demo$production, old_fscores$ability) 
cor.test(d_demo$production, old_fscores$ability) # .91

d_demo <- d_demo %>% left_join(old_fscores)

d_demo %>% ggplot(aes(x = age, y = ability, color=source)) + # color=sex
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), size = 1) +
  #geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
  geom_jitter(alpha = .2) + theme_bw() + xlab("Age (months)") + ylab("Ability")

plot(old_coefs$d, new_coefs$d)
cor(old_coefs$d, new_coefs$d) # .96
plot(old_coefs$a1, new_coefs$a1)  
cor(old_coefs$a1, new_coefs$a1) # .995
```


## Model comparison.

Compared to the Rasch model, the 2PL model fits better and is preferred by both AIC and BIC.

```{r, anovas, include=F}

# with d prior is better (also test a1 prior? norm(.2, 1))
#mc_p <- get_anova_table(mod_2pl, mod_2pl_pr, c("2PL", "2PL prior"))

mc1 <- get_anova_table(mod_1pl, mod_2pl, c("Rasch", "2PL"))
mc2 <- get_anova_table(mod_2pl, mod_3pl, c("2PL", "3PL"))
```

```{r, echo=F}
kable(mc1, digits=2, 
      caption="Comparison of Rasch and 2PL models.") %>%
      html_table_width(c(60, 80, 80, 80, 50))
```


The 2PL is favored over the 3PL model by both AIC and BIC.


```{r, echo=F}
kable(mc2, digits=2, 
      caption="Comparison of 2PL and 3PL models.") %>%
      html_table_width(c(60, 80, 80, 80, 50))
```



The 2PL is preferred over both the Rasch (1PL) model and the 3PL model, so we do the rest of our analyses using the 2PL model as the basis for the CAT.
Next we look for linear dependencies (LD) among the items, and also check for ill-fitting items. We will remove any items that show both strong LD and poor fit.


# Item bank

## Examine Linear Dependencies

```{r, linear-ind, echo=F, eval=RUN_MODELS}
#res = residuals(mod_2pl, type = 'JSI') 

res = residuals(mod_2pl, type = 'LD') # upper diag are standardized residuals in the form of signed Cramers V coefficients
# hist(unlist(res[upper.tri(res)]))
save(file=here("data/LD.Rds"), "res")
```

```{r, linear-dep-tab, echo=F, caption="Items showing moderate LD with more than 10 other items."}
load(here("data/LD.Rds"))

# no association = abs(V) < .1 no association, .3 is moderate, and .5+ is strong
#hiLDvio = get_LD_violations(res, assoc_str=.6) 3 words: item_3, item_8, item_12

hiLDvio = get_LD_violations(res, assoc_str=.5)
hiLDvio_words = coefs_2pl[which(hiLDvio>10),]$definition # 404 words... (was 8)
# "アイタ..いたい."  "オイチィ..おいしい."  "ニャンニャン..ネコ." "ブーブー..車."  "ワンワン..犬."      
# "クック.靴."   "ないない.片づけ."  "ネンネ"  

modLDvio = get_LD_violations(res)
modLDvio_words = coefs_2pl[which(modLDvio>10),]$definition # 700 words have moderate LD (!)

# ToDo: try just removing the few items with many violations (e.g., 1 item has 2 -- not bad!),
# and then re-fit the model

kable(matrix(coefs_2pl[which(hiLDvio>10),]$definition, ncol=2)) %>% 
  html_table_width(rep(150, 2))
```

We examined each item for pairwise linear dependencies (LD) with other items using $\chi^{2}$ (Chen & Thissen, 1997), and found that `r length(hiLDvio_words)` items show strong LD (Cramer's $V \geq 0.5$). 

## Ill-fitting items

Our next goal is to determine if all items should be included in the item bank. Items that have very bad properties should probably be dropped. We will prune any ill-fitting items ($\chi^{2}$ $p<.001$) from the full 2PL model that *also* showed strong LD.

```{r, item-fit-2pl, echo=F, eval=RUN_MODELS}
# use the 2PL fitted to all items
itfit2pl <- itemfit(mod_2pl, fit_stats = "X2") # need to use MAP estimates?
# had to switch from S_X2 because it can't handle missing data (all WG subjects)

itfit2pl_x2 <- itemfit(mod_2pl, fit_stats = 'X2*', Theta=fscores_2pl, return.tables = F) # 
# 'X2*' : Stone's (2000) fit statistics that require parametric bootstrapping
# 'PV_Q1*' : Chalmers and Ng's (2017) plausible-value variant of the Q1 statistic that uses parametric bootstrapping to obtain a suitable empirical distribution
save(file=here("data/jp_ws_wg_2pl_itemfits.Rds"), "itfit2pl") #  "itfit2pl_x2"
```

```{r, echo=F}
load(here("data/jp_ws_wg_2pl_itemfits.Rds"))
bad_items2pl = which(itfit2pl$p.X2 < .01) # 0

#med_rmsea = median(itfit2pl_x2$RMSEA.X2_star_scaled, na.rm=T) # .01
#length(which(itfit2pl$RMSEA.X2_star_scaled > .02))

# 0 items showing strong LD and poor fit
bad_ld_fit = intersect(which(hiLDvio > 0), bad_items2pl)
```

`r length(bad_items2pl)` items did not fit well in the full 2PL model <!-- , and these items are shown below. -->

```{r, echo=F}
kable(matrix(itfit2pl$item[bad_items2pl], ncol=4)) %>% 
  html_table_width(rep(150, 4))
```

<!-- Now we re-fit the 2PL model without the XXX items showing strong LD *and* poor fit removed.
(Note: should look removed/remaining items by category, difficulty, discrimination, and RMSEA.) -->

```{r, refit-pruned-2pl, echo=F, eval=RUN_MODELS}
#mod_2pl <- mirt(d_mat[,-bad_ld_fit], 1, itemtype='2PL',
#                technical=list(NCYCLES=3000))
pruned_nit = ncol(d_mat)-length(bad_ld_fit)
mod_string = paste0('G = 1-',pruned_nit,',
                LBOUND = (1-',pruned_nit,', a1, 0),
                PRIOR = (1-',pruned_nit,', d, norm, 0, 3)')
mod <- mirt.model(mod_string)
mod_2pl = mirt(data = d_mat[,-bad_ld_fit], model = mod, itemtype="2PL", 
                        method="QMCEM", verbose=TRUE, 
                        technical=list(NCYCLES=3000))

coefs_2pl <- as_tibble(coef(mod_2pl, simplify = TRUE)$items) %>%
  mutate(definition = rownames(coef(mod_2pl, simplify = TRUE)$items))
fscores_2pl <- tibble(data_id = rownames(d_mat), 
                      ability = fscores(mod_2pl, method = "MAP")[,1])

save(file = here("data/jp_ws_wg_mod_2pl_nobad.Rds"), "mod_2pl","fscores_2pl", "coefs_2pl")
# ToDo: should look at itemfits again and see if new items are bad

# load WS and merge by item_id
items <- read_csv("WS_WG_items_combined.csv")

coefs_2pl <- coefs_2pl %>% rename(item_id = definition) %>%
  left_join(items %>% rename(item_id = WS, definition = definition_ja3) %>% select(item_id, definition, definition_en, category_en)) 

# output parameters for catsim (MIRT uses slope-intercept, does catsim need other format?)
# discrimination, difficulty, guessing, upper asymptote?
write.csv(coefs_2pl, file="JP_production_2PL_params_slopeint.csv", row.names = F)
# ?mirt::coef 

coefs_2pl_irt <- as_tibble(coef(mod_2pl, simplify = TRUE, IRTpars=TRUE)$items) %>%
  mutate(WS = rownames(coef(mod_2pl, simplify = TRUE)$items)) %>%
  left_join(items %>% select(WS, definition_ja3)) %>% 
  rename(item_id = WS, definition = definition_ja3)
write.csv(coefs_2pl_irt, file="JP_production_2PL_params.csv", row.names = F)

#anova(mod_2pl, mod_3pl)
```

```{r, echo=F, eval=F}
# NOT EVALUATED FOR JAPANESE SINCE NO PRUNING NEEDED
load(here("data/jp_ws_wg_mod_2pl_nobad.Rds"))
```

```{r check-fit-ld-pruned-model, echo=F, eval=RUN_MODELS}
# are there a ton more ill-fitting items now? 
# has LD decreased after pruning?

itfit2pl <- itemfit(mod_2pl, fit_stats = "X2") 

itfit2pl[which(itfit2pl$p.X2<.001),] 

res = residuals(mod_2pl, type = 'LD') 
# hist(unlist(res[upper.tri(res)]))
#hist(res[1,])
hiLDvio = get_LD_violations(res, assoc_str=.5)
hiLDvio_words = coefs_2pl[which(hiLDvio>10),]$definition 

#intersect(which(hiLDvio>0), bad_items2pl) 

save(file=here("data/jp_pruned_LD_and_itemfit.Rds"), "res", "itfit2pl")
```



### Plot Item Parameters

Next, we examine the coefficients of the 2PL model. 
Items that are estimated to be very easy or very difficult are highlighted, as well as those at the extremes of discrimination (a1).

```{r, fig.width=8, fig.height=6, echo=F, warning=F}
# load WS and merge by item_id
items <- read_csv("WS_WG_items_combined.csv")

coefs_2pl <- coefs_2pl %>% rename(item_id = definition) %>%
  left_join(items %>% rename(item_id = WS, definition = definition_ja3) %>% select(item_id, definition, definition_en, category_en)) 

ggplot(coefs_2pl, 
       aes(x = a1, y = d)) + 
  geom_point(alpha=.7) + 
  ggrepel::geom_label_repel(data = filter(coefs_2pl, 
                                          a1 < 2.0 | d > 5.0 | 
                                          a1 > 5.0 | d < -3.0), 
                            aes(label = definition), family = "HiraKakuProN-W3") + 
  theme_bw()
```



Next, we will run simulated CATs on the data from the `r nrow(d_mat)` 12-36 month-olds. 
However, since many of these participants' data are from the CDI:WG form, there are many missing responses (compared to the CDI:WS). 
In order to run the simulated CATs, we impute the missing data using the participants' estimated ability and the 2PL model.
Overall, `r round(100*sum(is.na(d_mat))/length(d_mat),1)`% of the data was missing, and will be imputed.

```{r, impute-missing-data, echo=F, warning=F}
# To run the CAT simulations, we need to impute the missing responses
set.seed(123)
fs <- fscores(mod_2pl)
d_mat_imp = imputeMissing(mod_2pl, Theta=fs) # fscores_2pl has just ability, not discrim
```


# CAT Simulations

```{r, CAT, echo=F, eval = RUN_MODELS}
# parallelized
cores = detectCores()
cl = makeCluster(cores[1]-1) # don't overload computer
registerDoParallel(cl)

t25 = doCAT(d_mat_imp, mod_2pl, max_items=25, min_SEM=.1)
t50 = doCAT(d_mat_imp, mod_2pl, max_items=50, min_SEM=.1)
t75 = doCAT(d_mat_imp, mod_2pl, max_items=75, min_SEM=.1)
t100 = doCAT(d_mat_imp, mod_2pl, max_items=100, min_SEM=.1)
t200 = doCAT(d_mat_imp, mod_2pl, max_items=200, min_SEM=.1)
t300 = doCAT(d_mat_imp, mod_2pl, max_items=300, min_SEM=.1)
t400 = doCAT(d_mat_imp, mod_2pl, max_items=400, min_SEM=.1)
save("t25", "t50", "t75", "t100", 
     "t200", "t300", "t400", 
     file=here("data/jp_ws_wg_CAT_2pl.Rds"))

stopCluster(cl)
```

```{r, fixed-length-CAT, echo=F, eval = RUN_MODELS}
cores = detectCores()
cl = makeCluster(cores[1]-1) # don't overload computer
registerDoParallel(cl)

f25 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=25)
f50 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=50)
f75 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=75)
f100 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=100)
f200 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=200)
f300 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=300)
f400 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=400)
save("f25", "f50", "f75", "f100", 
     "f200", "f300", "f400", 
     file=here("data/jp_ws_wg_CAT_fixed_length.Rds"))
```

```{r, random-fixed-length, echo=F, eval = RUN_MODELS}

r25 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=25, criteria='random', start_item='random')
r50 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=50, criteria='random', start_item='random')
r75 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=75, criteria='random', start_item='random')
r100 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=100, criteria='random', start_item='random')
r200 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=200, criteria='random', start_item='random')
r300 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=300, criteria='random', start_item='random') # run these
r400 = doCAT_fixed_length(d_mat_imp, mod_2pl, min_items=400, criteria='random', start_item='random') # run these
save("r25", "r50", "r75", "r100", 
     "r200", "r300", "r400", 
     file=here("data/jp_ws_wg_random_test_fixed_length.Rds"))

stopCluster(cl)
```


For each wordbank subject, we simulate a CAT using a maximum of 25, 50, 100, 200, 300, or 400 items, with the termination criterion that it reach an estimated SEM of .1. 
For each of these simulations, we examine 1) which items were never used, 2) the median and mean number of items used, 3) the correlation of ability scores estimated from the CAT and from the full CDI, and 4) the mean standard error of the CATs. 

```{r, analyze-cat-sims, echo=F}
load(here("data/jp_ws_wg_CAT_2pl.Rds"))
load(here("data/jp_ws_wg_CAT_fixed_length.Rds"))
load(here("data/jp_ws_wg_random_test_fixed_length.Rds"))


cat_tab = summarize_CAT(t25, d_mat_imp)
cat_tab = rbind(cat_tab, summarize_CAT(t50, d_mat_imp))
cat_tab = rbind(cat_tab, summarize_CAT(t75, d_mat_imp))
cat_tab = rbind(cat_tab, summarize_CAT(t100, d_mat_imp))
cat_tab = rbind(cat_tab, summarize_CAT(t200, d_mat_imp))
cat_tab = rbind(cat_tab, summarize_CAT(t300, d_mat_imp))
cat_tab = rbind(cat_tab, summarize_CAT(t400, d_mat_imp))
cat_tab = data.frame(cat_tab)
names(cat_tab) = c("Maximum Qs", "Median Qs Asked", "Mean Qs Asked", "r with full CDI", "Mean SE", "Reliability", "Items Never Used")


kable(cat_tab, digits=3, caption="CAT simulations with 2PL model compared to full CDI.") %>%
  html_table_width(c(150,190,190,190, 150, 150, 200))

```


Finally, following Makransky et al. (2016), we run a series of fixed-length CAT simulations and again compare the thetas from these CATs to the ability estimates from the full CDI. The results are quite good even for 25- and 50-item tests, but note that we add a comparison to tests of randomly-selected questions (per subject), and find that ability estimates from these tests are also strongly correlated with thetas from the full CDI. The mean standard error of the random tests shows more of a difference.

```{r, fixed-length-CAT-table, echo=F}
fcat_tab = summarize_CAT(f25, d_mat_imp)
fcat_tab = rbind(fcat_tab, summarize_CAT(f50, d_mat_imp))
fcat_tab = rbind(fcat_tab, summarize_CAT(f75, d_mat_imp))
fcat_tab = rbind(fcat_tab, summarize_CAT(f100, d_mat_imp))
fcat_tab = rbind(fcat_tab, summarize_CAT(f200, d_mat_imp))
fcat_tab = rbind(fcat_tab, summarize_CAT(f300, d_mat_imp))
fcat_tab = rbind(fcat_tab, summarize_CAT(f400, d_mat_imp))
fcat_tab = data.frame(fcat_tab)
names(fcat_tab) = c("Test Length", "Median Qs Asked", "Mean Qs Asked", # drop Median / Mean
                    "r with full CDI", "Mean SE", "Reliability", "Items Never Used")
fcat_tab[,c("Median Qs Asked", "Mean Qs Asked")] = NULL
rand_baseline = c(cor(r25$parms$thetaCAT, fscores_2pl$ability),
                  cor(r50$parms$thetaCAT, fscores_2pl$ability),
                  cor(r75$parms$thetaCAT, fscores_2pl$ability),
                  cor(r100$parms$thetaCAT, fscores_2pl$ability),
                  cor(r200$parms$thetaCAT, fscores_2pl$ability),
                  cor(r300$parms$thetaCAT, fscores_2pl$ability), 
                  cor(r400$parms$thetaCAT, fscores_2pl$ability)) 

rand_SE = c(mean(r25$parms$CAT_SE), mean(r50$parms$CAT_SE), mean(r75$parms$CAT_SE),
            mean(r100$parms$CAT_SE), mean(r200$parms$CAT_SE), mean(r300$parms$CAT_SE), mean(r400$parms$CAT_SE))
fcat_tab[,"Random Test r with full CDI"] = rand_baseline
fcat_tab[,"Random Test Mean SE"] = rand_SE
kable(fcat_tab, digits=3, caption="Fixed-length CAT simulations with 2PL model compared to full CDI.") %>%
  html_table_width(c(130,160,130,130, 200, 200, 200))
```

```{r, echo=F}
d_demo$age_group = cut(d_demo$age, breaks=seq(11,38,3), right=F, include.lowest = T)
n_age = table(d_demo$age_group)
```

### Preferred CAT Settings

Testing with a minimum of 25 items, a maximum of 50, and termination at SE = .1, and ML scoring.
First we'll do it using the MI start item, and then we'll try choosing an age-based starting item per subject (based on mean theta for each age).


```{r, age-based-starting-item, echo=F}
# mean theta per age (mos)
#d_demo$data_id = as.character(d_demo$data_id)
age_startits <- d_demo %>% #left_join(fscores_2pl, by='data_id') %>%
  mutate(ability = fscores_2pl$ability) %>%
  group_by(age_mos) %>%
  summarise(theta = mean(ability), sd = sd(ability), n=n()) %>%
  rename(age = age_mos)
  

# find item with maximum information at given theta value

age_startits$definition = as.character(NA)
age_startits$index = as.integer(NA)
age_startits$item_info = as.numeric(NA)
```

```{r, age-based-starting-item2, echo=F}
# find item with max iteminfo at that theta
for(i in 1:nrow(age_startits)) {
  mi_it = get_max_info_item(mod_2pl, age_startits[i,]$theta-.5) # -.5 to make items easy
  age_startits[i,]$definition = as.character(coefs_2pl[mi_it$item,]$definition)
  age_startits[i,]$index = mi_it$item
  age_startits[i,]$item_info = mi_it$info # at average theta for age
}

start_item_subj = rep(NA, nrow(d_demo))
for(i in 1:nrow(d_demo)) {
  start_item_subj[i] = subset(age_startits, age==d_demo[i,]$age_mos)$index
}
write.csv(age_startits, file="JP_production_start_items_by_age.csv", row.names = F)
```


```{r, preferred-CATs, echo=F, eval=RUN_MODELS}
min25_max50_ML = preferredCAT(d_mat_imp, method='ML')
min25_max50_MAP = preferredCAT(d_mat_imp, method='MAP')
#get_cor_by_age(d_demo, min25_max50_ML) 

min25_max50_ML_age = preferredCAT(d_mat_imp, method='ML', start_item=start_item_subj)
min25_max50_MAP_age = preferredCAT(d_mat_imp, method='MAP', start_item=start_item_subj)
save(file="data/jp_preferredCAT.Rds", "min25_max50_ML", "min25_max50_MAP",
     "min25_max50_ML_age", "min25_max50_MAP_age")
```

We select a starting item with a difficulty just below the average ability (theta) for each age (in months). 
The mean theta per age is shown below, along with the selected starting item.

```{r, age-based-starting-items-table, echo=F}
kable(age_startits, digits=2) |>
  html_table_width(c(80,100,100,100, 120, 90, 100))
```


```{r, preferredCAT-table, echo=F}
load("data/jp_preferredCAT.Rds")
pref_tab = summarize_CAT(min25_max50_ML, d_mat_imp)
pref_tab = rbind(pref_tab, summarize_CAT(min25_max50_MAP, d_mat_imp))
pref_tab = rbind(pref_tab, summarize_CAT(min25_max50_ML_age, d_mat_imp))
pref_tab = rbind(pref_tab, summarize_CAT(min25_max50_MAP_age, d_mat_imp))
pref_tab = data.frame(pref_tab)
pref_tab$cond = c("ML / MI", "MAP / MI", "ML / age-based", "MAP / age-based")
names(pref_tab) = c("Scoring / Start Item", "Median Qs Asked", "Mean Qs Asked", "r with full CDI", "Mean SE", "Reliability", "Items Never Used")

# for Aaron
ml_vs_map = data.frame(cbind(MAPtheta = min25_max50_MAP$parms$thetaCAT, MLtheta = min25_max50_ML$parms$thetaCAT))
ggplot(aes(x=MAPtheta, y=MLtheta), data=ml_vs_map) + geom_point(alpha=.1) + theme_minimal()

# Bland-Altman plot
#ml_vs_map$Mean = with(ml_vs_map, (MAPtheta + MLtheta) / 2)
#ml_vs_map$MAPminusML =  with(ml_vs_map, MAPtheta - MLtheta)
#ml_vs_map$age_group = d_demo$age_group
#ggplot(aes(x=Mean, y=MAPminusML, color=age_group), data=ml_vs_map) + 
#  geom_point(alpha=.2) + theme_minimal() + xlab("Mean Theta") + ylab("MAP theta - ML theta")


kable(pref_tab, digits=3, caption="CAT simulations with min=25, max=50, stopping at SE=0.1.") %>%
  html_table_width(c(200,190,190,190, 150, 150, 200))
```


## Age analysis

Does the CAT show systematic errors with children of different ages? The table below shows correlations between ability estimates from the full CDI compared to the estimated ability from each fixed-length CAT split by age (`r n_age[1]` 11-13 month-olds, `r n_age[2]` 14-16 mos, `r n_age[3]` 17-19 mos, `r n_age[4]` 20-22 mos, `r n_age[5]` 23-25 mos, `r n_age[6]` 26-28 mos, `r n_age[7]` 29-31 mos, `r n_age[8]` 32-35 mos, and `r n_age[9]` 35-38 mos). This is comparable to Table 3 of Makransky et al. (2016), and the correlations here are consistently high across age groups.

```{r, cat-cor-age, echo=F}
d_demo$ability = fscores_2pl$ability


at1 = get_cor_by_age(d_demo, f25) 
at2 = get_cor_by_age(d_demo, f50) 
at3 = get_cor_by_age(d_demo, f75)
at4 = get_cor_by_age(d_demo, f100)
at5 = get_cor_by_age(d_demo, f200)
at6 = get_cor_by_age(d_demo, f300)
at7 = get_cor_by_age(d_demo, f400)

age_table = rbind(c(25, at1$r), 
                  c(50, at2$r),  
                  c(75, at3$r), 
                  c(100, at4$r), 
                  c(200, at5$r), 
                  c(300, at6$r), 
                  c(400, at7$r))
age_table = age_table[,1:10]
colnames(age_table) = c("Test Length", paste(names(n_age),"mos"))

kable(age_table, digits=3, caption="Correlation between fixed-length CAT ability estimates and the full CDI.") %>%
  html_table_width(rep(120, ncol(age_table)))
```

We further look at the correlations with age using the preferred CAT settings (min_items=25, max_items=50, stopping at SE=.15).

```{r, echo=F}
r1 = get_cor_by_age(d_demo, min25_max50_ML)
r2 = get_cor_by_age(d_demo, min25_max50_MAP)
r3 = get_cor_by_age(d_demo, min25_max50_ML_age)
r4 = get_cor_by_age(d_demo, min25_max50_MAP_age)

age_tabp = rbind(c("ML / MI", round(r1$r, 3)), 
                  c("MAP / MI", round(r2$r, 3)), 
                  c("ML / age-based", round(r3$r, 3)), 
                  c("MAP / age-based", round(r4$r, 3)))
age_tabp = age_tabp[,1:10]
colnames(age_tabp) = c("Scoring / Start Item", paste(names(n_age),"mos"))

kable(age_tabp, digits=3, caption="Correlation between the preferred CAT's ability estimates and the full CDI.") %>%
  html_table_width(c(220, rep(120, ncol(age_tabp)-1)))
```


Below we show the distribution of ability (theta) from the 2PL model by age.

```{r}
d_demo |> filter(is.na(age_group))
```


```{r, theta-by-age, echo=F, fig.width=6, fig.height=4}
# Overlaid histograms
d_demo |> filter(!is.na(age_group)) |>
  ggplot(d_demo, aes(x=ability, fill=age_group)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity") + theme_classic() +
  xlab("Ability (Theta) Estimates from 2PL") +
  labs(fill="Age Group") + ylab("Number of Participants")

# Density plots with semi-transparent fill
#ggplot(d_demo, aes(x=ability, fill=age_group)) + geom_density(alpha=.3) + theme_classic()
```


## Ability analysis

Finally, we ask whether the fixed-length CATs work well for children of different abilities.
Below are scatterplots that show the standard error estimates vs. estimated ability (theta) for each child on the different simulated fixed-length CATs. The 25-item CAT shows some visible distortion, but the 50-item CAT is already quite smooth, and the 75-item CAT indistinguishable from the 300- or 400-item CATs. 
Based on these plots and the above tables we may recommend that users adopt a 50-item CAT using the 2PL parameters, but suggest that they may want to administer a full CDI if the participant's estimated theta from the CAT is <-0.5 or >2 (where the SE from CAT starts to exceed 0.1).

```{r, echo=F, fig.width=10, fig.height=10, caption="Scatterplots showing the standard error estimates for each child conditionally on the basis of estimated ability level for the different simulated fixed-length CATs."}

eg25 <- f25$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("25-item CAT")
eg50 <- f50$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("50-item CAT")

eg75 <- f75$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("75-item CAT")
eg100 <- f100$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("100-item CAT")

eg300 <- f300$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("300-item CAT")
eg400 <- f400$parms %>% ggplot(aes(x=thetaCAT, y=CAT_SE)) + geom_point(alpha=.2) +
    xlab("Theta Estimate from CAT") + ylab("SE from CAT") + theme_bw() + ggtitle("400-item CAT")

grid.arrange(eg25, eg50, 
             eg75, eg100,
             eg300, eg400, nrow = 3)

```



## Item selection for item bank

```{r, include=F}
f50items = sort(table(f50$all_items))
f75items = sort(table(f75$all_items))
f100items = sort(table(f100$all_items))
cat_freq = rbind(data.frame(test=50, item=names(f50items), N=as.vector(f50items)),
                 data.frame(test=75, item=names(f75items), N=as.vector(f75items)),
                 data.frame(test=100, item=names(f100items), N=as.vector(f100items)))  

cat_freq$prop = cat_freq$N / nrow(d_mat_imp)
cat_freq$bin = cut(cat_freq$prop, breaks=seq(0,1,.1))

#wcat <- eng_ws %>% group_by(definition, lexical_class, category) %>%
#  filter(!is.na(category)) %>% 
#  summarise(produces = mean(produces, na.rm=T)) %>%
#  select(definition, lexical_class, category, produces)

pop_items = subset(cat_freq, test==50 & prop>.5) # item_159, item_294, item_234, item_382
pop_items$item = as.numeric(as.character(pop_items$item))
pop_items$definition = colnames(d_mat_imp)[pop_items$item]
#wcat[as.vector(pop_items$item),] # no longer indexed correctly due to pruning
```

Of the `r ncol(d_mat_imp)` pruned CDI:WS items, `r length(f50items)` were selected on one or more administrations of the fixed-length 50-item CATs simulated from the wordbank data.
Which items were most frequently selected for the fixed-length 50-item CAT? 
Shown in the table below, only `r nrow(pop_items)` items were selected on more at least 50% of the tests.

```{r, frequent-cat-items, echo=F}
pop50_tab = tibble(Item=pop_items$definition, Proportion=round(pop_items$prop,2)) %>% arrange(-Proportion)
kable(pop50_tab, digits=3, caption="Items chosen on at least 50% of the 50-item CATs.") %>%
  html_table_width(rep(150, 2))
```


Below we show the overall distribution of how many of the `r ncol(d_mat_imp)` pruned CDI:WS items were selected on what percent of the CATs of varying length (50, 75, or 100 items). Note that we do not include in the graph the number of items that were never selected on each test: `r ncol(d_mat_imp)-length(f50items)` items never selected on the 50-item test, `r ncol(d_mat_imp)-length(f75items)` items on the 75-item test, and `r ncol(d_mat_imp)-length(f100items)` items never selected on the 100-item test. The longer the test, the less skewed the distribution, but even on the 100-item CAT most of the appearing items are selected less than a third of the time.

```{r, cat-item-distro, echo=F, fig.width=6, fig.height=4, caption="Distribution of use of items in the 50-, 75-, and 100-item variable-length CATs (not including items that were never selected)."}
cat_freq$Length = as.factor(cat_freq$test)

ggplot(cat_freq, aes(x=bin, group=test, fill=Length)) + geom_bar(position=position_dodge()) + theme_bw() + ylab("Number of Items")
```

```{r, echo=F}
never_selected = setdiff(1:ncol(d_mat_imp), unique(t300$all_items)) # t400 = 39 103 237 472
```


Below we show the `r length(never_selected)` items from the pruned CDI:WS that were never selected on the maximum 300-item CAT. 
<!-- Somewhat surprisingly, this set has no overlap with the set of low-information items identified earlier. -->


```{r, echo=F}
#intersect(bad_items, colnames(d_mat)[never_selected])
kable(matrix(colnames(d_mat_imp)[never_selected], ncol=4)) %>%
  html_table_width(rep(150, 4))
```


What about the items that are most selected across all of the CATs (25-400-item)? Here are the top 50:

```{r, echo=F}
all_items = sort(table(c(t25$all_items, t50$all_items, t75$all_items, t100$all_items, t200$all_items, t300$all_items, t400$all_items)), decr=T)

kable(matrix(colnames(d_mat_imp)[as.numeric(names(all_items[1:50]))], ncol=5)) %>%
  html_table_width(rep(150, 7))
```


These are predominantly...


## Example CAT

```{r, include=F}
# generate response patterns
set.seed(123)
pattern0 <- generate_pattern(mod_2pl, Theta = 0) # sum(pattern0) = 562
pattern1 <- generate_pattern(mod_2pl, Theta = 1) # sum(pattern1) = 642

choices <- matrix(c(0,1), nrow(coefs_2pl), 2, byrow = TRUE)
questions <- coefs_2pl$definition
df <- data.frame(Question=questions, 
                 Option = choices, 
                 Type = 'radio', stringsAsFactors = FALSE)

preferred_design = list(min_items = 25,
                        max_items = 50, 
                        min_SEM = 0.10)

# start_item = 'MI',
# index 14 = au.revoir (start item for 12-14 mos)
res0 <- mirtCAT(df, mod_2pl, local_pattern=pattern0, criteria = 'MI', start_item = 14, method = 'ML', 
               design = preferred_design) 

# index 616 = construire (start item for 24 mos)
res1 <- mirtCAT(df, mod_2pl, local_pattern=pattern1, criteria = 'MI', start_item = 616, method = 'ML', 
               design = preferred_design) 

seq0 = tibble(item = coefs_2pl[res0$items_answered,]$definition,
         response = res0$scored_responses[res0$items_answered],
         theta = res0$thetas_history[-1])

seq1 = tibble(item = coefs_2pl[res1$items_answered,]$definition,
         response = res1$scored_responses[res1$items_answered],
         theta = res1$thetas_history[-1])

# FixMe
write.csv(seq0, file="jp_CAT_test_theta0_au.revoir.csv", row.names = F) # Theta_1 = 0.01
write.csv(seq1, file="jp_CAT_test_theta1_construire.csv", row.names = F) # Theta_1 = 1.20 
```


We now show an example CAT for two simulated participants, one with ability (theta) = 0, and one with theta = 1. The CAT gives a minimum of 25 questions and terminates either when SEM=0.1 or when 50 items is reached.
The theta estimates over the test for each participant is shown below, with selected item indices on the x axis. The theta=0 participant (left) answered `r length(res0$items_answered)` questions, and the theta=1 participant (right) answered `r length(res1$items_answered)`. The final estimated theta for the theta=0 participant was `r round(as.vector(res0$thetas),3)`, and for the theta=1 participant was `r round(as.vector(res1$thetas),3)`. The package `mirtCAT` can be directly used to simply generate a web interface (Shiny app) that allows such CATs to be run on real participants, as well as the simulations we have conducted here.


```{r, echo=F, fig.width=10, fig.height=4}
#summary(res1)
grid.arrange(plot(res0), plot(res1), nrow=1)
```


```{r, include=F}
#CATdesign <- mirtCAT(df, mod_2pl, criteria = 'MI', design_elements = TRUE)
#findNextItem(CATdesign)
#CATdesign2 <- updateDesign(CATdesign, items = c(592), responses = c(0))
#findNextItem(CATdesign2)

```


### References


Makransky, G., Dale, P. S., Havmose, P. and Bleses, D. (2016). An Item Response Theory–Based, Computerized Adaptive Testing Version of the MacArthur–Bates Communicative Development Inventory: Words & Sentences (CDI:WS). *Journal of Speech, Language, and Hearing Research*. 59(2), pp. 281-289.



