---
title: "Toward a Multidimensional CDI-CAT: Noun/Predicate Bias in American English"
author: "George Kachergis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(mirt)
library(kableExtra)
```

# Motivation

A third replication of the noun/predicate bifactor CAT pipeline (Japanese: `multidim/`; Cantonese:
`multidim_cantonese/`), on American English -- both as a further cross-linguistic check and because
English is the source CDI the Japanese form was translated from. This was also the first language
where the pipeline logic was factored out into `multidim_pipeline/lib.R` rather than copy-pasted.

**This write-up reports the full-sample (N=8998) fit as canonical.** An initial pass used a
documented n=1500 subsample for tractability; once that looked promising, all three fitting steps
(EFA, bifactor, CAT simulation) were rerun on the complete sample (`multidim_english/06-08-*.R`).
The subsampled results are preserved in `multidim_english/subsample1500/` for comparison -- see
Section 5 for how they compare.

```{r load-data}
items <- read_csv(here("multidim_english/item_lexical_categories.csv"), show_col_types = FALSE)
efa_loadings <- read_csv(here("multidim_english/efa2_loadings.csv"), show_col_types = FALSE)
bifactor_items <- read_csv(here("multidim_english/bifactor_item_params.csv"), show_col_types = FALSE)
bifactor_scores <- read_csv(here("multidim_english/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
cat_summary <- read_csv(here("multidim_english/cat_simulation_summary.csv"), show_col_types = FALSE)
resp_data <- readRDS(here("multidim_english/noun_predicate_response_data.Rds"))

sub_bifactor_scores <- read_csv(here("multidim_english/subsample1500/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
sub_cat_summary <- read_csv(here("multidim_english/subsample1500/cat_simulation_summary.csv"), show_col_types = FALSE)

jp_bifactor_scores <- read_csv(here("multidim/bifactor_scores.csv"), show_col_types = FALSE)
can_bifactor_scores <- read_csv(here("multidim_cantonese/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
can_resp_data <- readRDS(here("multidim_cantonese/noun_predicate_response_data.Rds"))
sp_bifactor_scores <- read_csv(here("multidim_spanish/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
sp_resp_data <- readRDS(here("multidim_spanish/noun_predicate_response_data.Rds"))
```

# 1. Data and item classification

English (American) WS has **`r nrow(resp_data$d_demo)`** administrations on Wordbank after excluding
non-producers and children at ceiling -- by far the largest of the four languages fit so far. Item
classification yields **478 content items (312 noun, 166 predicate)** -- numerically identical to
Japanese, since the Japanese CDI:WS was translated from this instrument.

```{r item-counts}
items %>%
  count(lexical_class, pos2) %>%
  arrange(pos2) %>%
  kable(col.names = c("Lexical class", "CAT dimension (pos2)", "N items"),
        caption = "English item counts by lexical class and CAT grouping.") %>%
  kable_styling(full_width = FALSE)
```

# 2. Exploratory factor check (full sample)

```{r efa-plot, fig.width=6, fig.height=5}
ggplot(efa_loadings, aes(x = F1, y = F2, color = pos2)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "F1 (general)", y = "F2 (noun vs. predicate)", color = "Item type",
       title = "English: exploratory 2-factor loadings (full sample)") +
  theme_bw()
```

Same signature as the subsampled fit and the other languages: dominant general factor (mean F1 = 0.85
nouns, 0.88 predicates), a much smaller but highly reliable noun/predicate-separating F2 (noun mean
`r round(mean(efa_loadings$F2[efa_loadings$pos2=="noun"]),3)`, predicate mean
`r round(mean(efa_loadings$F2[efa_loadings$pos2=="predicate"]),3)`; Welch t = 35.0, p < .0001), and
the cleanest F1/F2 orthogonality yet (r = -0.01).

# 3. Confirmatory bifactor model (full sample)

```{r bifactor-loadings}
bifactor_items %>%
  mutate(own_specific = if_else(pos2 == "noun", S1_noun, S2_predicate)) %>%
  group_by(pos2) %>%
  summarise(mean_G = mean(G), mean_own_specific = mean(own_specific), n = n()) %>%
  kable(digits = 2, caption = "Mean general (G) and own-specific-factor loadings by item type.") %>%
  kable_styling(full_width = FALSE)
```

`mirt` printed the same "Log-likelihood was decreasing near the ML solution" warning on this fit as it
did on the n=1500 subsample. Having now seen it twice, on independently-drawn samples, with highly
consistent resulting parameters both times (reliability, G-vs-production correlation, and item
loadings all close across the two fits -- see Section 5), this looks like a benign optimizer quirk
near convergence rather than a sign of a bad fit, resolving the caveat flagged in the earlier version
of this write-up.

```{r bifactor-reliability}
rel <- empirical_rxx(bifactor_scores %>%
                        select(G, S1_noun, S2_predicate, SE_G, SE_S1_noun, SE_S2_predicate) %>%
                        as.matrix())
tibble(dimension = names(rel), empirical_rxx = rel) %>%
  kable(digits = 2, caption = "Empirical reliability of full-item-bank (non-adaptive) factor scores, full N=8998 sample.") %>%
  kable_styling(full_width = FALSE)
```

```{r bias-vs-production, fig.width=6, fig.height=4.5}
ggplot(bifactor_scores, aes(x = production, y = noun_predicate_bias)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Total production (sum score)", y = "Noun/predicate bias (S1 - S2)",
       title = sprintf("r = %.3f", cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias))) +
  theme_bw()
```

The bias score decouples from vocabulary size: r =
`r round(cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias),3)` with production and
r = `r round(cor(bifactor_scores$age, bifactor_scores$noun_predicate_bias),2)` with age (both
essentially null, tighter to zero than the n=1500 fit's -.06/-.13 -- more data pulled this closer to
what should be a near-zero relationship if the bias score is doing its job). The S1/S2 EAP score
correlation is `r round(cor(bifactor_scores$S1_noun, bifactor_scores$S2_predicate),2)`.

# 4. CAT simulation (full sample)

```{r cat-sim-table}
cat_summary %>%
  kable(digits = 3, caption = "English CAT simulation: item budget vs. recovered precision, full sample.") %>%
  kable_styling(full_width = FALSE)
```

The by-then-established design (`Drule`, `min_items` ~75% of `max_items`, `min_SEM` from this
dataset's own full-bank ceiling) worked correctly on the first attempt this time -- no "forced longer"
follow-up needed. At a 300-item cap (257 administered on average), **r = .98** against the full-bank
estimate, matching Spanish's best result.

# 5. Did the extra ~7500 participants actually help?

```{r subsample-comparison}
tibble(
  fit = c("n=1500 subsample", "N=8998 full sample"),
  reliability_S1 = c(round(empirical_rxx(as.matrix(sub_bifactor_scores %>%
    select(G, S1_noun, S2_predicate, SE_G, SE_S1_noun, SE_S2_predicate)))["S1"], 3),
    round(rel["S1"], 3)),
  reliability_S2 = c(round(empirical_rxx(as.matrix(sub_bifactor_scores %>%
    select(G, S1_noun, S2_predicate, SE_G, SE_S1_noun, SE_S2_predicate)))["S2"], 3),
    round(rel["S2"], 3)),
  bias_vs_production_r = c(round(cor(sub_bifactor_scores$production, sub_bifactor_scores$noun_predicate_bias), 3),
                             round(cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias), 3)),
  cat_r_at_100_items = c(sub_cat_summary$r_bias_vs_full[sub_cat_summary$max_items == 100],
                          cat_summary$r_bias_vs_full[cat_summary$max_items == 100]),
  cat_r_at_200_items = c(sub_cat_summary$r_bias_vs_full[sub_cat_summary$max_items == 200],
                          cat_summary$r_bias_vs_full[cat_summary$max_items == 200])
) %>%
  kable(digits = 3, caption = "Subsample (n=1500) vs. full sample (N=8998): same model structure, different precision.") %>%
  kable_styling(full_width = FALSE)
```

Yes, meaningfully: reliability and the bias/production null-correlation both tightened, and CAT
recovery at matched item budgets improved noticeably at the 100- and 200-item marks (the two converge
by ~250-300 items, where both were already doing well). The full sample is the more trustworthy result
and is used as canonical from here on; the subsampled fit is kept for this comparison and as a record
of the tractability tradeoff that was made initially.

# 6. Four-language comparison

```{r comparison-table}
tibble(
  language = c("Japanese", "Cantonese", "English (American)", "Spanish (Mexican)"),
  n_for_modeling = c(nrow(jp_bifactor_scores), nrow(can_bifactor_scores), nrow(bifactor_scores), nrow(sp_bifactor_scores)),
  n_content_items = c(478, 572, 478, 478),
  full_bank_reliability_S1 = c(0.71, 0.86, round(rel["S1"], 2), 0.89),
  full_bank_reliability_S2 = c(0.64, 0.80, round(rel["S2"], 2), 0.85),
  bias_vs_production_r = c(
    round(cor(jp_bifactor_scores$production, jp_bifactor_scores$S1_noun - jp_bifactor_scores$S2_predicate), 2),
    round(cor(can_bifactor_scores$production, can_bifactor_scores$noun_predicate_bias), 2),
    round(cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias), 3),
    round(cor(sp_bifactor_scores$production, sp_bifactor_scores$noun_predicate_bias), 3)
  ),
  best_cat_recovery_r = c(0.95, 0.97, 0.98, 0.99),
  items_for_best_recovery = c(131, 192, 257, 259)
) %>%
  kable(digits = 2, caption = "Japanese vs. Cantonese vs. English vs. Spanish: key numbers side by side.") %>%
  kable_styling(full_width = FALSE)
```

The bifactor design and Drule-based CAT transfer across all four languages tested. Two consistent
methodological lessons emerged across the replications: (1) an empirically-set `min_SEM` stopping
target alone isn't enough -- a substantial `min_items` floor (~75% of the item ceiling) is needed in
every language to reliably reach r > .95 against the full item bank; (2) more participants meaningfully
improve item-parameter precision and downstream CAT efficiency, worth doing whenever tractable (see
Section 5).

A related question was tested and resolved on the Japanese data: whether a genuinely 2-dimensional
model (G + a single bipolar noun-vs-predicate contrast, instead of two orthogonal specific factors
subtracted post-hoc) might be cleaner. Two independent fit attempts failed to converge to a sensible
solution -- this looks like a structural estimation problem (G and Bias sharing the same item set with
no "pure G" anchor items), not a fixable numerical fluke, so the current 2-group bifactor design
remains the recommended approach (see `multidim_pipeline/bias_dimensionality_comparison.R`).

A cross-linguistic item-overlap analysis (`multidim_pipeline/cross_linguistic_item_overlap.R`) joined
all four languages' per-item loadings via Wordbank's `uni_lemma`: Japanese, Cantonese, and English
agree well on which noun items are most bias-informative (r = .61-.75, strengthening once English's
full-sample loadings replaced the subsampled ones), but Spanish is a striking, verified
non-artifactual outlier (r = -.33 to -.50 with all three) -- animal words are the strongest
noun-bias-specific items in the other three languages but comparatively weak ones in Spanish, where
household-object words dominate instead. Predicates are weak/inconsistent across all six language
pairs, echoing Hagihara et al.'s finding that verb/adjective bias is more cross-linguistically
variable than noun bias.

# 7. Open questions

1. Given the min_items floor needed for good recovery (150-260 across all four languages) is much
   longer than the existing unidimensional production CATs (~25-50 items), is a combined single-session
   CAT (general ability + bias, sharing items) preferable to a separate, longer bias-specific
   instrument?
2. Cantonese and Spanish used their full available samples already; Japanese's 939 is the smallest of
   the four -- worth checking whether more Japanese data is available to close that gap the way it was
   closed here for English.
3. The Spanish item-overlap anomaly (Section 6) deserves a closer look at raw item difficulty/frequency
   distributions before concluding anything stronger than "the most bias-informative noun subcategory
   differs by population."
