---
title: "Toward a Multidimensional CDI-CAT: Noun/Predicate Bias in American English"
author: "George Kachergis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(mirt)
library(kableExtra)
```

# Motivation

A third replication of the noun/predicate bifactor CAT pipeline (Japanese: `multidim/`; Cantonese:
`multidim_cantonese/`), this time on American English -- both as a further cross-linguistic check and
because English is the source CDI the Japanese form was translated from, making it a useful reference
point. This is also the first language where the pipeline logic itself was factored out into
`multidim_pipeline/lib.R` rather than copy-pasted, since a third from-scratch implementation was no
longer justified.

```{r load-data}
items <- read_csv(here("multidim_english/item_lexical_categories.csv"), show_col_types = FALSE)
efa_loadings <- read_csv(here("multidim_english/efa2_loadings.csv"), show_col_types = FALSE)
bifactor_items <- read_csv(here("multidim_english/bifactor_item_params.csv"), show_col_types = FALSE)
bifactor_scores <- read_csv(here("multidim_english/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
cat_summary <- read_csv(here("multidim_english/cat_simulation_summary.csv"), show_col_types = FALSE)
resp_data <- readRDS(here("multidim_english/noun_predicate_response_data.Rds"))
d_sub <- readRDS(here("multidim_english/modeling_subsample.Rds"))

jp_bifactor_scores <- read_csv(here("multidim/bifactor_scores.csv"), show_col_types = FALSE)
jp_items <- read_csv(here("multidim/item_lexical_categories.csv"), show_col_types = FALSE)
can_bifactor_scores <- read_csv(here("multidim_cantonese/bifactor_scores.csv"), show_col_types = FALSE) %>%
  mutate(noun_predicate_bias = S1_noun - S2_predicate)
can_items <- read_csv(here("multidim_cantonese/item_lexical_categories.csv"), show_col_types = FALSE)
```

# 1. Data, item classification, and a scale decision

English (American) WS has **`r nrow(resp_data$d_demo)`** administrations on Wordbank after excluding
non-producers and children at ceiling -- 7-9x more than Japanese (939) or Cantonese (1205). Item
classification uses the same category &rarr; class mapping (cross-validated against Wordbank's own
`lexical_category`, which again collapses verbs+adjectives into "predicates"), yielding **478 content
items (312 noun, 166 predicate)** -- numerically identical to Japanese, unsurprising since the
Japanese CDI:WS was translated from this instrument.

```{r item-counts}
items %>%
  count(lexical_class, pos2) %>%
  arrange(pos2) %>%
  kable(col.names = c("Lexical class", "CAT dimension (pos2)", "N items"),
        caption = "English item counts by lexical class and CAT grouping.") %>%
  kable_styling(full_width = FALSE)
```

**Scale decision**: fitting the EFA/bifactor models on the full ~9000-person sample would plausibly
take hours (Cantonese, at 572 items x 1205 people, already took 30/12.6 min for EFA/bifactor; English
at 478 items x 8998 people is ~7.5x more people). Rather than let that run unattended for an uncertain
number of hours, a fixed random subsample of **n = `r nrow(d_sub$d_mat_content)`** administrations
(seed = 1234) was drawn and used consistently for the EFA, the confirmatory bifactor fit, and the CAT
simulation's full-item-bank comparison (all three must use the same people). This is a documented
tractability decision -- refitting on the full sample is a natural follow-up once the subsampled
results are judged worth the extra compute, and it would only improve precision, not change the
qualitative picture below.

# 2. Exploratory factor check

```{r efa-plot, fig.width=6, fig.height=5}
ggplot(efa_loadings, aes(x = F1, y = F2, color = pos2)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "F1 (general)", y = "F2 (noun vs. predicate)", color = "Item type",
       title = "English: exploratory 2-factor loadings") +
  theme_bw()
```

Third replication of the same signature: dominant general factor (mean F1 = 0.85 nouns, 0.89
predicates), a much smaller but highly reliable noun/predicate-separating F2 (noun mean
`r round(mean(efa_loadings$F2[efa_loadings$pos2=="noun"]),3)`, predicate mean
`r round(mean(efa_loadings$F2[efa_loadings$pos2=="predicate"]),3)`; Welch t = 30.6, p < .0001), and the
cleanest F1/F2 orthogonality of the three languages so far (r = -0.002, vs. Japanese's -0.015 and
Cantonese's 0.234).

# 3. Confirmatory bifactor model

```{r bifactor-loadings}
bifactor_items %>%
  mutate(own_specific = if_else(pos2 == "noun", S1_noun, S2_predicate)) %>%
  group_by(pos2) %>%
  summarise(mean_G = mean(G), mean_own_specific = mean(own_specific), n = n()) %>%
  kable(digits = 2, caption = "Mean general (G) and own-specific-factor loadings by item type.") %>%
  kable_styling(full_width = FALSE)
```

One caveat worth flagging before trusting these numbers: `mirt` printed a warning during this fit
("Log-likelihood was decreasing near the ML solution. EM method may be unstable"). Item parameters
were checked for anything degenerate (extreme loadings, near-zero/boundary values) and none were
found; the CAT simulation below (Section 4) also ultimately recovers the full-bank bias estimate at
r = .98 given enough items, which would be hard to explain if the fit had actually landed somewhere
badly wrong. Still, a from-different-starting-values refit would be worth doing before treating these
English item parameters as final, since this is the only one of the three language fits that raised
this warning.

```{r bifactor-reliability}
rel <- empirical_rxx(bifactor_scores %>%
                        select(G, S1_noun, S2_predicate, SE_G, SE_S1_noun, SE_S2_predicate) %>%
                        as.matrix())
tibble(dimension = names(rel), empirical_rxx = rel) %>%
  kable(digits = 2, caption = "Empirical reliability of full-item-bank (non-adaptive) factor scores, on the n=1500 subsample.") %>%
  kable_styling(full_width = FALSE)
```

```{r bias-vs-production, fig.width=6, fig.height=4.5}
ggplot(bifactor_scores, aes(x = production, y = noun_predicate_bias)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Total production (sum score)", y = "Noun/predicate bias (S1 - S2)",
       title = sprintf("r = %.2f", cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias))) +
  theme_bw()
```

The bias score decouples from vocabulary size here too -- in fact more cleanly than either other
language: r = `r round(cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias),2)` with
production (Japanese: -.16, Cantonese: -.11) and r =
`r round(cor(bifactor_scores$age, bifactor_scores$noun_predicate_bias),2)` with age. The S1/S2 EAP
score correlation is `r round(cor(bifactor_scores$S1_noun, bifactor_scores$S2_predicate),2)` -- between
Japanese's .48 and Cantonese's .72.

# 4. CAT simulation

`Drule` again matched-or-beat `Wrule` on a subsample check and was used throughout. `min_SEM` was set
adaptively from this dataset's own full-bank SE ceiling (`r round(median(bifactor_scores$SE_S1_noun),2)`
/ `r round(median(bifactor_scores$SE_S2_predicate),2)` for S1/S2) rather than reused from another
language -- the Cantonese replication had already shown that reusing another language's target badly
under-shoots.

```{r cat-sim-table}
cat_summary %>%
  kable(digits = 3, caption = "English CAT simulation: item budget vs. recovered precision.") %>%
  kable_styling(full_width = FALSE)
```

Even with the adaptively-set target, recovery at a 200-item cap (mean 132 items administered) was
weaker than either other language (r = .92, vs. Japanese's .95 and Cantonese's .97 at similar or
smaller item counts). Forcing more items (`min_items = 200`, tightened `min_SEM = c(0.3,0.3,0.3)`,
capped at 250) resolved this the same way it did for Cantonese: mean items rose to 245, and
**r(CAT bias, full-bank bias) reached .98** -- the best of the three languages once given enough items.
So this was again a stopping-rule calibration issue, not evidence that the flagged convergence warning
compromised the model.

# 5. Three-language comparison

```{r comparison-table}
tibble(
  language = c("Japanese", "Cantonese", "English (American)"),
  n_for_modeling = c(nrow(jp_bifactor_scores), nrow(can_bifactor_scores), nrow(bifactor_scores)),
  n_available = c(nrow(jp_bifactor_scores), nrow(can_bifactor_scores), nrow(resp_data$d_demo)),
  n_content_items = c(478, 572, 478),
  full_bank_reliability_S1 = c(0.71, 0.86, 0.80),
  full_bank_reliability_S2 = c(0.64, 0.80, 0.74),
  bias_vs_production_r = c(
    round(cor(jp_bifactor_scores$production, jp_bifactor_scores$S1_noun - jp_bifactor_scores$S2_predicate), 2),
    round(cor(can_bifactor_scores$production, can_bifactor_scores$noun_predicate_bias), 2),
    round(cor(bifactor_scores$production, bifactor_scores$noun_predicate_bias), 2)
  ),
  best_cat_recovery_r = c(0.95, 0.97, 0.98),
  items_for_best_recovery = c(131, 192, 245)
) %>%
  kable(digits = 2, caption = "Japanese vs. Cantonese vs. English: key numbers side by side.") %>%
  kable_styling(full_width = FALSE)
```

The bifactor design and Drule-based CAT transfer across all three languages tested. One consistent
methodological lesson emerged from doing this three times: **an empirically-set `min_SEM` stopping
target still isn't enough on its own** -- Cantonese and English both showed the CAT stopping well
short of the item counts actually needed for high-fidelity bias recovery, even with a target
calibrated to that dataset's own precision ceiling. A substantial `min_items` floor (on the order of
150-250, well above what's needed for general ability alone) currently seems necessary in all three
languages to reliably reach r > .95 against the full item bank.

# 6. Open questions

1. **Bifactor refit stability**: worth a from-different-starting-values refit of the English model
   given the convergence warning, even though downstream checks suggest it isn't consequential.
2. **Full-sample refit**: the n=1500 subsample was a tractability call; with ~9000 administrations
   available, a full-sample (or larger subsample) refit would tighten these estimates further and is
   easy to justify given how promising the subsampled results are.
3. Given the min_items floor needed for good recovery (150-250) is much longer than the existing
   unidimensional production CATs (~25-50 items), is a combined single-session CAT (general ability +
   bias, sharing items) preferable to a separate, longer bias-specific instrument, across all three
   languages?
4. The `multidim_pipeline/lib.R` refactor means adding a fourth language is now mostly: pull data,
   check the category mapping needs no additions, decide on a subsample size if N is large, and run.
   Worth deciding whether further languages are wanted before over-investing in generalizing further
   (e.g. automatic WG+WS reconciliation, which was skipped here since English's WS-only sample was
   already large enough).
