---
title: "Cross-linguistic item overlap for the noun/predicate bias dimension"
author: "George Kachergis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(kableExtra)
```

# Question

Are the same concepts (Wordbank's `uni_lemma` -- a language-independent identifier linking
translation-equivalent CDI items, e.g. "dog" links English *dog*, Japanese *inu*, Cantonese *gau2*,
Spanish *perro*) the most informative items for the noun-vs-predicate bias dimension across languages,
or is item-level informativeness mostly language-specific? Each language's confirmatory bifactor model
(`multidim*/bifactor_item_params.csv`) gives every item a loading on its own specific factor (S1 for
nouns, S2 for predicates) -- a direct, per-item measure of how much that item discriminates
noun-leaning from predicate-leaning children. This compares those loadings across all four languages
fit so far: Japanese, Cantonese, English (American), Spanish (Mexican).

```{r load-data}
wide <- read_csv(here("multidim_pipeline/cross_linguistic_item_loadings_wide.csv"), show_col_types = FALSE)
pair_cors <- read_csv(here("multidim_pipeline/cross_linguistic_pairwise_correlations.csv"), show_col_types = FALSE)
top_overlap <- read_csv(here("multidim_pipeline/cross_linguistic_top_item_overlap.csv"), show_col_types = FALSE)
all_items <- read_csv(here("multidim_pipeline/cross_linguistic_item_loadings.csv"), show_col_types = FALSE)

lang_labels <- c(multidim = "Japanese", multidim_cantonese = "Cantonese",
                  multidim_english = "English", multidim_spanish = "Spanish")
pair_cors <- pair_cors %>% mutate(lang1 = recode(lang1, !!!lang_labels), lang2 = recode(lang2, !!!lang_labels))
top_overlap <- top_overlap %>% mutate(lang1 = recode(lang1, !!!lang_labels), lang2 = recode(lang2, !!!lang_labels))
```

# 1. Pairwise correlation of item loadings

For every pair of languages, the correlation (across items sharing a `uni_lemma`) between how strongly
each item loads on its own specific factor in each language.

```{r cor-table}
pair_cors %>%
  arrange(pos2, lang1, lang2) %>%
  kable(digits = 2, col.names = c("Language 1", "Language 2", "Item type", "N shared items", "r")) %>%
  kable_styling(full_width = FALSE) %>%
  row_spec(which(grepl("Spanish", paste(pair_cors$lang1, pair_cors$lang2))), background = "#fff3cd")
```

Three languages (Japanese, Cantonese, English) agree well with each other on nouns (r = .61-.71) and
much less on predicates (r = .15-.73, inconsistent). **Every pair involving Spanish is negative for
nouns** (r = -.42 to -.50) and mostly negative for predicates too (r = -.41 to -.44, except Japanese-Spanish
predicates at +.29). This is highlighted rather than averaged away, because it turned out to be a real
and informative pattern, not noise (Section 3).

# 2. Top-30 most-discriminating-item overlap vs. chance

For each language pair, how many of the 30 most-discriminating items (by loading magnitude) share a
`uni_lemma`, compared to the number expected if item informativeness were independent across languages.

```{r overlap-table}
top_overlap %>%
  arrange(pos2, lang1, lang2) %>%
  kable(digits = 1, col.names = c("Language 1", "Language 2", "Item type", "Top N", "Shared-item pool",
                                    "Observed overlap", "Chance-expected overlap")) %>%
  kable_styling(full_width = FALSE)
```

Japanese-Cantonese-English noun overlap runs 2-3x above chance. Every pair involving Spanish's nouns
shows **zero** observed overlap against an expected ~4-5 by chance -- not just "no enrichment," but
below what random assignment would produce, which is the signature of systematically different items
being informative, not merely noisier estimates.

# 3. Is this a sign-convention artifact?

Bifactor specific-factor loadings don't have a globally fixed sign (unlike a single bipolar dimension,
each specific factor's orientation is whatever the model happened to converge to), so a first hypothesis
was that Spanish's S1_noun dimension simply came out sign-flipped relative to the other three languages.
Checked directly: flipping Spanish's sign does turn the raw correlation positive, but doesn't recover a
sensible "same items, just inverted" pattern -- rather than mirroring which items *are* top-loading,
Spanish's ranking is a genuinely different ordering.

```{r animal-household-check}
animals <- c("tiger", "turtle", "fish (animal)", "frog", "bird", "giraffe", "lion", "penguin",
             "chick", "horse", "monkey", "bear", "squirrel", "pig")
household <- c("stove", "money", "iron (object)", "bag", "broom", "scissors", "soap", "sweater",
               "comb (object)", "basket")

check <- all_items %>%
  filter(uni_lemma %in% c(animals, household), pos2 == "noun") %>%
  mutate(category = if_else(uni_lemma %in% animals, "animal", "household"),
         language = recode(language, !!!lang_labels)) %>%
  group_by(language, category) %>%
  summarise(mean_own_specific_loading = round(mean(own_specific_loading), 2), n = n(), .groups = "drop") %>%
  pivot_wider(names_from = category, values_from = c(mean_own_specific_loading, n))

check %>%
  kable(caption = "Mean noun-specific (S1) loading for the same animal vs. household-object items, by language.") %>%
  kable_styling(full_width = FALSE)
```

This is the actual mechanism: in Japanese/Cantonese/English, **animal words are the strongest
noun-specific-bias indicators**; in Spanish, **animal words are comparatively weak noun-bias indicators
and household-object words are the strongest**. Checking Spanish's own G vs. S1 loadings for these
items directly confirms why: Spanish animal words have high G loadings (2.0-3.2, near the top of the
noun G distribution) but comparatively low S1 loadings (0.18-1.24) -- i.e. in the Spanish (Mexican)
CDI sample, knowing an animal word is a strong marker of *overall* vocabulary size but doesn't carry
much *extra* information about noun-vs-predicate bias specifically, whereas the reverse is true for
household objects. This isn't a model-fitting artifact -- Spanish's bifactor fit is the cleanest of
the four languages by every other diagnostic (highest reliability, cleanest convergence, best CAT
recovery) -- it's a genuine difference in which noun subcategories carry bias-specific signal in this
population, plausibly reflecting how early and uniformly animal words are acquired in Mexican Spanish
CDI norms specifically (making them excellent general-vocabulary markers but poor bias-specific
markers), though confirming that explanation would need a look at the raw item difficulty/frequency
distributions rather than resting on this loading pattern alone.

# 4. Summary

- **Nouns are not universally cross-linguistically transferable at the item level** -- the original
  3-language result (Japanese/Cantonese/English agreeing at r=.6-.7) looked like it might generalize,
  but Spanish breaks it cleanly and systematically, not through noise but through a real, checkable
  difference in which noun subcategory (animals vs. household objects) carries bias-specific signal.
- **Predicates remain the weaker, more inconsistent case** across all six language pairs, consistent
  with Hagihara et al.'s finding that verb/adjective bias is more cross-linguistically variable than
  noun bias -- though with only 4 languages and a mix of positive and negative correlations, this is
  suggestive rather than conclusive.
- **Practical implication for item bank design**: a cross-linguistically "universal" short-list of
  maximally-informative noun items (e.g., for a rapid cross-language screening tool) is not supported
  by this data -- the most useful items for the bias dimension appear to be genuinely
  population/language-specific, at least between these four languages.
